---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rray

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/DavisVaughan/rray.svg?branch=master)](https://travis-ci.org/DavisVaughan/rray)

## Introduction

The goal of rray is to provide stricter matrix and array classes in R, similar in spirit to `tibble`.

It also supports _broadcasting_ semantics for matrices and arrays using the `xtensor` C++ library. This greatly extends the flexibility of calculations you can perform with them. This does require C++14.

## Installation

You can install from Github with:

```{r, eval = FALSE}
devtools::install_github("DavisVaughan/rray")
```

## mtrx and rray

```{r}
library(rray)
```

The rray package introduces two new classes. An `rray`, which is an N-dimensional object and essentially a stricter array, and a `mtrx`, which is a subclass of an `rray` that restricts the object to only two dimensions. 

rray consistently takes the approach that all inputs have at least two dimensions, with scalars and vectors being treated as 1 column matrices.

The easiest way to create a `mtrx` is to use the constructor. It takes in (potentially named) vectors and transforms them into the columns of the matrix.

```{r}
mtrx_ex <- mtrx(b = 1:2, a = 3:4)

mtrx_ex
```

mtrices are stricter than base R's matrix objects in two main ways:

* You must be explicit about the dimension when selecting columns (i.e. `x[,1]` not `x[1]`).
* Dimensions are never dropped.

```{r, error = TRUE}
# first column
mtrx_ex[,1]

# not this
mtrx_ex[1]

# individual elements
mtrx_ex[1, "a"]
```

These rules also apply to `rray` objects. You can create an `rray` from a vector, matrix, or array.

```{r}
# 2 rows, 3 columns, 2 "deep" (being the 3rd dimension)
x_rray <- rray(1, dim = c(1, 3, 2))

x_rray
```

```{r}
# first col
x_rray[,1]

# first row of each dimension
x_rray[1,]

# select the first element in the 3rd dimension
x_rray[,,1]
```

## Broadcasting

Broadcasting, the concept of recycling dimensions of one object to match the 
dimensions of another, is one of the core differences between operations with
rray and base R.

It allows you to do powerful operations such as:

```{r}
x_rray 

x_rray + 1:3
```

To learn how this is working, see the broadcasting vignette, `vignette("broadcasting", "rray")`.

## Motivation

The motivation for rray started with the inability to do this one operation with
base R.

```{r, error=TRUE}
x <- matrix(1:5)
y <- matrix(1:10, ncol = 2)

x 

y

x + y
```

As shown earlier, broadcasting solves this problem. I believe that broadcasting is a natural expression of intent when working with arrays, and is fully _shape-stable_ in the fact that the dimensions of the outputs are always fully defined by the dimensions of the inputs, so I feel that its adoption in R would be worthwhile.

A second motivation is the handling of dimension names in base R. While consistent, 
it lacks in expressibility that I feel should be there, and results can be surprising.

```{r}
x <- mtrx(a = 1:2, b = 3:4)
y <- mtrx(5:6, 7:8, row_names = c("r1", "r2"))

x_mat <- as.matrix(x)
y_mat <- as.matrix(y)

x_mat

y_mat

# Dimension names of x_mat are used
x_mat + y_mat

# Dimension names of y_mat are used
y_mat + x_mat
```

With rray, dimension names can come from both `x` and `y`. The rules for 
determining dimension names in rray are spelled out in `?rray_dim_names_common`.

```{r}
x + y
```

## Alternatives

I don't know of any other R packages that attempt to implement broadcasting
semantics.

The original motivation for this package, and even for `xtensor`, is the excellent
Python library, `numpy`. As far as I know, it has the original implementation of
broadcasting, and is a core library that a huge number of others are built on 
top of.
