---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rray

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/DavisVaughan/rray.svg?branch=master)](https://travis-ci.org/DavisVaughan/rray)
[![Codecov test coverage](https://codecov.io/gh/DavisVaughan/rray/branch/master/graph/badge.svg)](https://codecov.io/gh/DavisVaughan/rray?branch=master)
<!-- badges: end -->

## Introduction

rray (said: "r-ray") is an array manipulation library for R. It has three main goals:

1) To provide an rray class that implements stricter matrices and arrays than base R, similar in spirit to tibble. `vignette("the-rray", package = "rray")`.

2) To support _broadcasting_ semantics throughout the package, allowing for more flexible and intuitive array operations than are possible with base R. `vignette("broadcasting", package = "rray")`.

3) To provide a consistent, powerful toolkit for array based manipulation, usable by rray objects and (importantly!) by base R matrices and arrays as well. `vignette("toolkit", package = "rray")`.

View the vignette for each goal to learn more about how to use rray.

```{r}
library(rray)
```

## What can it do?

In short, rray tries to make array manipulation in R more intuitive by combining the ideas of broadcasting and knowing when to _not_ drop dimensions. This results in operations such as:

```{r}
x <- rray(1:6, dim = c(3, 2))

# Compute proportions along the 1st dimension
x / rray_sum(x, axes = 1)

# Equivalent base R syntax
sweep(x, 2, apply(x, 2, sum), "/")
```

These concepts are baked into every part of rray, and show up in other functions such as `rray_bind()`. Using broadcasting, `rray_bind()` can bind arrays together in ways that base R cannot with the native `cbind()` and `rbind()` functions.

```{r, error=TRUE}
a <- array(c(1, 2), dim = c(2, 1))
b <- array(c(3, 4), dim = c(1, 2))

a

b

# Error
cbind(a, b)

# `a` is broadcast to have dimensions: (2, 2)
rray_bind(a, b, axis = 1)

# Error
rbind(a, b)

# `b` is broadcast to have dimensions: (2, 2)
rray_bind(a, b, axis = 2)
```

## Installation

You can install from Github with:

```{r, eval = FALSE}
devtools::install_github("DavisVaughan/rray")
```

## Acknowledgements

rray would not be possible without the underlying C++ library, [`xtensor`](https://github.com/QuantStack/xtensor). Additionally, rray uses a large amount of the infrastructure in [`vctrs`](https://github.com/r-lib/vctrs) to be as consistent and type stable as possible. 

## Alternatives

The Matrix package implements a small subset of column-wise broadcasting 
operations. rray fully supports broadcasting in all operations.

The original motivation for this package, and even for xtensor, is the excellent
Python library, numpy. As far as I know, it has the original implementation of
broadcasting, and is a core library that a huge number of others are built on 
top of.

In the past, the workhorse for flexibly binding arrays together has been the abind package. This package has been a great source of inspiration and has served as a nice benchmark for rray.

## Notes

Currently, rray does not handle missing values. This is coming, as the underlying library xtensor natively supports missing values, however a few upstream bugs are currently preventing rray from using those features.

rray will perform best on R 3.6.0 and above. It is able to take advantage of a few of the ALTREP features there, which result in less copies being made per function call.

