---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rray

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/DavisVaughan/rray.svg?branch=master)](https://travis-ci.org/DavisVaughan/rray)


The goal of `rray` is to provide stricter matrix and array classes in R, similar in spirit
to `tibble`.

It also supports broadcasting semantics for matrices and arrays using the
`xtensor` library. This greatly extends the flexibility of calculations you
can perform with matrices.

**This is still just an experiment.**

## Installation

You can install from Github with:

```{r, eval = FALSE}
devtools::install_github("DavisVaughan/rray")
```


## Creation

```{r}
library(rray)
```


Create `mtrx` objects by either coercing existing objects to `mtrx` or
by using the helpful constructor, `mtrx()`.

```{r}
# Coercion
mat <- matrix(1:10, nrow = 5, dimnames = list(NULL, c("a", "b")))

as_mtrx(mat)
```

The constructor takes multiple vectors in `...`, where each element is a column
of the matrix, possibly named.

```{r}
mtrx_ex <- mtrx(b = c(1, 2, 3, 4, 5), a = c(6, 7, 8, 9, 10))

mtrx_ex
```

mtrx objects have the same slicing semantics as tibbles. Dimensions are never
dropped.

```{r}
# first column
mtrx_ex[,1]

# first row
mtrx_ex[1,]

# second column by name
mtrx_ex["a"]

# individual elements
mtrx_ex[1, "a"]
```

Broadcast semantics are available through `xtensor`.

```
(5, 2)
(5, 1)
------
(5, 2)
```

```{r, error=TRUE}
col_a <- mtrx_ex["a"]
col_a

mtrx_ex + col_a

# Note that you cant do this in base R!
as_matrix(mtrx_ex) + as_matrix(col_a)
```

Outer sums and products are incredibly simple with this broadcast syntax.

```
(5, 1)
(1, 5)
------
(5, 5)
```


```{r}
col_a + t(col_a)

col_a * t(col_a)
```

It's not just restricted to matrices either. Let's create a 2x3x2 rray object.

```{r}
x_rray <- rray(c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12))
x_rray <- rray_reshape(x_rray, c(2, 3, 2))

x_rray
```

Subsetting rray objects is strict too.

```{r}
# first col
x_rray[,1]

# first row of each dimension
x_rray[1,]

# only the first 3rd dimension object
x_rray[,,1]
```

Broadcasting operations are more powerful here.

```{r}
x_rray[,1]
```

```
(2, 3, 2)
(2, 1, 2)
---------
(2, 3, 2)
```

```{r}
x_rray - x_rray[,1]
```

```
(2, 3, 2)
(1, 3, 2)
---------
(2, 3, 2)
```

```{r}
x_rray + x_rray[1,]
```

You can mix rray and mtrx objects, and more importantly you can mix both with
base R types.

```{r}
two_rows <- col_a[1:2,]
two_rows

# (2, 3, 2)
# (2, 1,  )
# ---------
# (2, 3, 2)
x_rray + two_rows
```

```{r}
# Divide by an R matrix
x_rray / matrix(5, nrow = 2, ncol = 3)

# Multiply an R "scalar"
x_rray / 2

# R vectors are treated as 1 column matrices
x_rray + c(10, 11)
```

You can manually broadcast with `rray_broadcast()`

```{r}
rray_broadcast(c(10, 11), c(2, 10))

# broadcasting a mtrx to 3 dimensions result in an rray
rray_broadcast(col_a, c(5, 3, 2))
```

There are a number of other functions in `xtensor` that would be great to expose.
One of them is `atan()` and it has been exposed already.

```{r}
atan(col_a)

atan(x_rray)
```

