---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rray

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/DavisVaughan/rray.svg?branch=master)](https://travis-ci.org/DavisVaughan/rray)

## Introduction

The goal of rray is to provide stricter matrix and array classes in R, similar in spirit to `tibble`.

It also supports _broadcasting_ semantics for matrices and arrays using the `xtensor` C++ library. This greatly extends the flexibility of calculations you can perform with them. This does require C++14.

## Installation

You can install from Github with:

```{r, eval = FALSE}
devtools::install_github("DavisVaughan/rray")
```

## The rray

```{r}
library(rray)
library(magrittr)
```

The rray package introduces a new array-like object, the `rray`. This is an N-dimensional object and essentially a stricter array.

rray allows the creation of 1D objects, which look similar to vectors, but function similar to 1 column matrices (in terms of how operations are broadcasted).

The easiest way to create an `rray` is to use the constructor. It takes an object which could be a vector, matrix, or array, and creates a new `rray` with optional reshaping.

```{r}
x_rray <- rray(1:4, c(2, 2)) %>%
  set_col_names(c("a", "b"))

x_rray
```

rrays are stricter than base R's rray objects in two main ways:

* You must be explicit about the dimension when selecting columns (i.e. `x[,1]` not `x[1]`).
* Dimensions are never dropped.

```{r, error = TRUE}
# first column
x_rray[,1]

# not this
x_rray[1]

# individual elements
x_rray[1, "a"]
```

It's easy to create 3D objects (and higher dimensions) as well.

```{r}
# 2 rows, 3 columns, 2 "deep" (being the 3rd dimension)
y_rray <- rray(1, dim = c(1, 3, 2))

y_rray
```

The same subsetting rules apply here, allowing for more intuitive subset operations.

```{r}
# first col
y_rray[,1]

# first row of each dimension
y_rray[1,]

# select the first element in the 3rd dimension
y_rray[,,1]
```

## Broadcasting

Broadcasting, the concept of recycling dimensions of one object to match the 
dimensions of another, is one of the core differences between operations with
rray and base R.

It allows you to do powerful operations such as:

```{r}
y_rray 

y_rray + 1:3
```

To learn how this is working, see the broadcasting vignette, `vignette("broadcasting", "rray")`.

## Motivation

The motivation for rray started with the inability to do this one operation with
base R.

```{r, error=TRUE}
x <- matrix(1:5)
y <- matrix(1:10, ncol = 2)

x 

y

x + y
```

As shown earlier, broadcasting solves this problem. I believe that broadcasting is a natural expression of intent when working with arrays, and is fully _shape-stable_ in the fact that the dimensions of the outputs are always fully defined by the dimensions of the inputs, so I feel that its adoption in R would be worthwhile.

A second motivation is the handling of dimension names in base R. While consistent, 
it lacks in expressibility that I feel should be there, and results can be surprising.

```{r}
z_rray <- rray(5:8, c(2, 2)) %>%
  set_row_names(c("r1", "r2"))

x_mat <- as.matrix(x_rray)
z_mat <- as.matrix(z_rray)

x_mat

z_mat

# Dimension names of x_mat are used
x_mat + z_mat

# Dimension names of z_mat are used
z_mat + x_mat
```

With rray, dimension names can come from both objects. The rules for 
determining dimension names in rray are spelled out in `?rray_dim_names_common`.

```{r}
x_rray + z_rray
```

## Alternatives

The `Matrix` package implements a small subset of column-wise broadcasting 
operations. `rray` fully supports broadcasting in all operations.

The original motivation for this package, and even for `xtensor`, is the excellent
Python library, `numpy`. As far as I know, it has the original implementation of
broadcasting, and is a core library that a huge number of others are built on 
top of.
