---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rray

[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/DavisVaughan/rray.svg?branch=master)](https://travis-ci.org/DavisVaughan/rray)

## Introduction

rray has three goals:

1) To provide an rray class that implements stricted matrices and arrays than base R, similar in spirit to the tibble package. `vignette("rray", package = "rray")`.

2) To support _broadcasting_ semantics throughout the package, allowing for more flexible and intuitive array operations than are possible with base R. `vignette("rray", package = "broadcasting")`.

3) To provide a consistent, powerful toolkit for array based manipulation.

View the vignette for each goal to learn more about how to use rray.

## Installation

You can install from Github with:

```{r, eval = FALSE}
devtools::install_github("DavisVaughan/rray")
```

## Acknowledgements

rray would not be possible without the underlying C++ library, [`xtensor`](https://github.com/QuantStack/xtensor). Additionally, rray uses a large amount of the new infrastructure in [`vctrs`](https://github.com/r-lib/vctrs) to be as consistent and type stable as possible. 

## Alternatives

The Matrix package implements a small subset of column-wise broadcasting 
operations. rray fully supports broadcasting in all operations.

The original motivation for this package, and even for `xtensor`, is the excellent
Python library, `numpy`. As far as I know, it has the original implementation of
broadcasting, and is a core library that a huge number of others are built on 
top of.
