---
title: "Broadcasting"
author: "Davis Vaughan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Broadcasting}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Broadcasting

```{r}
library(rray)
```

### Rules

Broadcasting is the concept of recycling dimensions of one object to match the dimensions of another so that some arithmetic operation can be applied.

It sounds complicated, but it's much easier to think about if you simply write down what is happening. There are only three rules:

1) Dimensions of size 1 are always recycled up to the other dimension you are comparing to.

2) Implicit dimensions are always appended to the right hand side, and are assumed to be 1's (this is actually different from how it is done in Python).

3) If two dimensions do not match, and neither are 1, an error is thrown.

These three rules also happen to match the "tidy" recycling rules.

### Example

As an example, consider a 3x2 matrix and a 3x1 matrix. With base R, you can't 
add these together.

```{r, error=TRUE}
x <- matrix(1:2, nrow = 3, ncol = 2, byrow = TRUE)
y <- matrix(1,   nrow = 3, ncol = 1)

x + y
```

With broadcasting semantics, the 3x1 matrix is "broadcasted" up to a 3x2 matrix,
and then the operation is applied. Essentially, the 1st column is copied into
the 2nd column.

```{r}
x_mtrx <- as_mtrx(x)
y_mtrx <- as_mtrx(y)

x_mtrx + y_mtrx
```

Internally, this is what is happening:

```{r}
y_bcasted <- rray_broadcast(y_mtrx, c(3, 2))

y_bcasted

x_mtrx + y_bcasted
```

Personally, I find it helpful to write it down:

```
(3, 2) <- a 3x2 matrix
(3, 1) <- combined with a 3x1
------
(3, 2) <- results in a 3x2
```

Broadcasting can be applied to higher dimensions as well. Let's try an example 
with an array and a matrix.

```{r}
x_rray <- rray(1, dim = c(3, 4, 2))

y_mtrx + x_rray
```

To understand this, let's again write it out:

```
(3, 1,  ) <- implicit 1 here
(3, 4, 2)
---------
(3, 4, 2)
```

Both the explicit 1 and the implicit 1 in the 3x1 matrix are broadcasted up
to match the other dimensions so that the 3x1 matrix becomes a 3x4x2 array.

```{r}
rray_broadcast(y_mtrx, c(3, 4, 2))
```

### Outer sums

Broadcasting makes certain calculations surprisingly simple. Consider this 
outer sum calculation:

```{r}
z <- as_mtrx(1:3)

z + t(z)
```

```
(3, 1)
(1, 3)
------
(3, 3)
```

It's this kind of powerful calculation that makes me think that R could greatly
benefit from the adoption of these semantics.
