---
title: "`rray`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{`rray`}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rray)
```

## Introduction

rray provides a new array-like class that alters some of the behavior with base R arrays that I find non-intuitive. In the same spirit as the tibble package, rray does less than base R, but works as consistently and intuitively as possible.

## Subsetting

Base R subsets matrices and arrays with a default of `drop = TRUE`, similar to data frames. This can result in problematic behavior, especially when programming around arrays, because a lot of mental work is required to ensure that functions don't accidentally subset matrices in a way that coerces them to vectors.

```{r}
x_mat <- matrix(1:6, ncol = 2)
x_mat

x_mat[1:2,]

x_mat[1,]

x_mat[,1:2]

x_mat[,1]
```

Generally, the fix for this is to use `drop = FALSE`.

```{r}
x_mat[1, , drop = FALSE]
```

But this requires careful thinking about how to subset arrays programatically, adding the right number of commas where required to ensure that dimensions aren't dropped. Consider pulling the first row of each dimension, and pay attention to the varying number of commas required to prevent a vector from being returned.

```{r}
x_3D <- array(1:12, c(2, 3, 2))
x_4D <- array(1:12, c(2, 3, 2, 1))

x_3D[1, , , drop = FALSE]

x_4D[1, , , , drop = FALSE]
```

rray takes a different approach, and never drops dimensions when subsetting. This approach actually frees up more intuitive subsetting syntax, which actually results in an error with base R.

```{r, error = TRUE}
x_3D_rray <- as_rray(x_3D)

x_3D_rray[1,]

x_3D[1,]
```

In this way, whether subsetting one element or multiple elements from a certain dimension, the result is an array with the same dimensionality as before. This is more predictable to program around.

```{r}
x_3D_rray[, 1]

x_3D_rray[, 1:2]
```

## Element Access

Base R allows you to access inner elements of an array by using `[` without any commas. Again, this dropping of dimensions feels inconsistent, and is rarely something that I've wanted to happen.

```{r}
x_3D[1:2]
```

rray disables this behavior, and instead throws an error. 

```{r, error = TRUE}
x_3D_rray[1:2]
```

Instead, either coerce to a vector then subset the inner elements, or first use the more explicit rray subsetting behavior to isolate the elements (keeping the dimensionality) and then call `as.vector()`. Alternatively, to keep an rray but drop dimensions of size `1`, use `rray_squeeze()`.

```{r}
# coerce then subset
as.vector(x_3D_rray)[1:2]

# subset
x_slice <- x_3D_rray[1:2, 1, 1]
x_slice

# then coerce
as.vector(x_slice)

# or drop dimensions of size 1
# to keep an rray
rray_squeeze(x_slice)
```

An argument could be made for the behavior of `[` with no commas to work like data frames, returning columns.

```{r}
df <- data.frame(x = 1:3, y = 4:6)

df[1]
```

This works fine for 2 dimensional data frames, where returning the columns can be justified as the most intuitive thing to do (especially for data analysis). But for arrays with 3 or more dimensions, the mental model of "rows" and "columns" is generally traded in for simply saying "the first dimension". In this world, there is no strong case for `[` with no commas to have a special behavior of always returning elements from an arbitrary 2nd dimension.

## Dimension Names

Base R has consistent dimension name handling rules, but they can be surprising because they often don't retain the maximum amount of information that it seems like they could.

```{r}
x_row_nms <- matrix(1:2, dimnames = list(c("r1", "r2")))
x_col_nms <- matrix(1:2, dimnames = list(NULL, c("c1")))

x_row_nms

x_col_nms

# names from x_row_nms are used
x_row_nms + x_col_nms

# names from x_col_nms are used
x_col_nms + x_row_nms
```

There is no reason that dimension names from both inputs can't be used here. rray has a different set of dimension name reconciliation rules that attempts to pull names from all inputs.

```{r}
x_row_nms_rray <- as_rray(x_row_nms)
x_col_nms_rray <- as_rray(x_col_nms)

x_row_nms_rray + x_col_nms_rray

x_col_nms_rray + x_row_nms_rray
```

This isn't to say that the order doesn't matter. If both inputs have row names, for example, the row names of the result come from the first input.

```{r}
x_col_row_nms_rray <- set_row_names(x_col_nms_rray, c("row1", "row2"))
x_col_row_nms_rray

x_row_nms_rray + x_col_row_nms_rray

x_col_row_nms_rray + x_row_nms_rray
```

This approach emphasizes maintaining the maximum amount of information possible. The underlying engine for dimension name handling is `rray_dim_names_common()`. Pass it multiple inputs, and it will return the common dimension names using rray's rules.

```{r}
rray_dim_names_common(x_row_nms, x_col_nms)
```

